Probabilistic Network Latency Verification

### 2023-02-22 paper/1_intro.tex
[Paper] Added new overview diagram
[Paper] Updated figures and polish

Over the last decade, there has been a lot of development in the area of network verification 
generally \cite{hsa}\cite{veriflow}.
Still, most of these verifiers are primarily fixated on verifying \textit{qualitative} properties 
-- the convergent functional behavior of a network under various failure scenarios -- 
such as reachability and loop detection, and not \textit{quantitative} properties 
related to performance, such as bandwidth and latency.
Compared to the verification of reachability-based property, verification of bandwidth and latency is 
inherently more challenging since it introduces more dimension to the problem.
In addition to reasoning about the existence of a flow, one needs to reason about the measured 
detail of said flow and model their interaction.

There has been a few notable works that push network verification concept towards this direction \cite{qarc}
\cite{ccac}.
These newer tools model a wide range of additional network behavior that might affect performance to answer a
specific question in mind, namely link-load violation and congestion control behavior.
However, their approaches so far has been focused on worst-case analysis, modeling and answering whether 
an unwanted event could happen or not, which requires further statistical interpretation in order to translate 
them to a percentage guarantee that appear on an SLA.

Moreover, none of these tools were designed to answer questions related to end-to-end latency SLA \cite{Verizon}.
Ideally, such a tool would answer the challenge of modeling various latency altering network behavior, 
such as traffic generation, load balancing scheme, and congestion control, and expresses the important 
statistical property of the resulting end-to-end latency.
Not just in terms of worst-case behaviors and tails, but also the means.

@@

Our evaluation shows that the verification framework, with the help of our optimization techniques, 
could accomplish the verification task on various WAN and datacenter topologies in the order of 
minutes?

@@

The code for this work is open-sourced at link. %TODO: change link
This work does not raise any ethical issues.

\begin{figure*}[h]
    \centering
    \includegraphics{../tikz/process}
    \caption{Overview of Tempus}
    \label{fig:process}
\end{figure*}

>>>>

Verification of network properties related to performance metric, like bandwidth and latency, is inherently 
more challenging than a classic reachabality-based property since it introduces more dimension to the problem.
In addition to reasoning about the existence of a flow, one needs to reason about the measured 
detail of said flow and model their interaction.

Recently, there has been a few notable works that push network verification concept towards this direction 
\cite{qarc}\cite{pita}\cite{ccac}.
These \textit{quantitative} verifier model a wide range of additional network behavior that might affect 
performance to answer a specific question in mind, such as link-load violation and congestion control behavior.
However, their approaches so far has been focused on worst-case analysis, modeling and answering whether 
an unwanted event could happen or not, which requires further statistical interpretation in order to translate 
them to a percentage guarantee that appear on an SLA.

Moreover, none of these tools were designed to answer questions related to end-to-end latency, 
which is one of many important property that needs to be made certain by network providers \cite{Verizon}.
Ideally, such a tool would answer the challenge of modeling various latency altering network behavior, 
such as traffic generation, load balancing scheme, and congestion control, and expresses the important 
statistical property of the resulting end-to-end latency.
This is done not just in terms of worst-case behaviors and tails, but also the means.

@@

Our evaluation shows that the verification framework, with the help of our optimization techniques, 
could accomplish the verification task on various WAN and datacenter topologies around 90\% faster than 
our unoptimized baseline.
Moreover, highly symmetrical topology, like Fat Tree, benefits from our optimization the most 
and could reduce the verification time up to 6 orders of magnitude.

@@

The code for this work is open-sourced on GitHub. %TODO: change link
This work does not raise any ethical issues.

\begin{figure*}[h]
    \centering
    \includegraphics[scale=0.5]{overall}
    \caption{Overview of Tempus}
    \label{fig:process}
\end{figure*}

### 2023-02-22 paper/2_overview.tex
[Paper] Polish
[Paper] Updated figures and polish

% \section{Overview}
\section{Latency Modeling Background}
In the context of verification, there are multiple approaches to model latency in a network.
We will briefly compare 4 possible methods that might be useful to attack this problem and 
and its limitation to illustrate the methodological challenge to accurately model latency.

\subsection{Simulation}
One of the available methods to reason about latency in a network is by simulating it 
using a network simulator (e.g. ns3, omnet).
On a high level, the tool will simulate the lifecycle of packets, created by a load generator, 
and the user could fail some network components and collect the packets' latency measurement 
to see how the perturbation affects the latency property in question.

@@

\subsection{Timed Automata}
Timed Automata is an extension to the classic Finite State Machine (FSM) in automata theory that 
adds a global clock to be used as transition constraints. 
The runtime will keep a set of global clock that monotonously increases and can be individually 
reset.
Using these clocks, we could construct the state machine that constraints each transition with 
a time limit. 
With this additional time constraints, we could verify properties in a similar way 
as traditional model checking methods (e.g. using CTL formulas).

@@

A popular Timed Automata modeling tool, UPPAAL, tried to alleviate this determinism problem by 
introducing Statistical Model Checking (SMC) on top of their Timed Automata model. 
The feature allows user to introduce random transition (with configurable weights) and distribution 
based transition, in order to probabilistically reason about the desired property and model 
random failures of network components.

@@

\subsection{Latency as Random Variable}
% Verification framework that is
% - Expressive (congestion control, load-balancing scheme, traffic generation -> probabilistic)
% - Performant (verification time that makes sense, proper-level of abstraction)
% - Practical (reasonably accurate while requiring minimal amount of information and assumption)

>>>>

% \section{Overview}
\section{Latency Modeling Background}
In the context of verification, there are multiple approaches to model latency in a network.
We will briefly compare 3 possible methods that might be useful to attack this problem and 
and its limitation to illustrate the methodological challenge to accurately model latency.

\subsection{Simulation}
One of the available methods to reason about latency in a network is by simulating it 
using a network simulator (e.g. ns-3 \cite{ns3}, OMNeT \cite{omnet}).
On a high level, the tool will simulate the lifecycle of packets, created by a load generator, 
and the user could fail some network components and collect the packets' latency measurement 
to see how the perturbation affects the latency property in question.

@@

\subsection{Timed Automata}
The other alternative is by using Timed Automata.
Timed Automata is an extension to the classic Finite State Machine (FSM) in automata theory where it 
adds a global clock to be used as transition constraints. 
The runtime will keep a set of global clock that monotonously increases and can be individually 
reset.
Using these clocks, we could construct the state machine that constraints each transition with 
a time limit. 
With this additional time constraints, we could verify properties in a similar way 
as traditional model checking methods (e.g. using CTL formulas).

@@

A popular Timed Automata modeling tool, UPPAAL \cite{uppaal}, tried to alleviate this determinism problem by 
introducing Statistical Model Checking (SMC) on top of their Timed Automata model. 
The feature allows user to introduce random transition (with configurable weights) and distribution 
based transition, in order to probabilistically reason about the desired property and model 
random failures of network components.

@@

\subsection{\tool}
% Verification framework that is
% - Expressive (congestion control, load-balancing scheme, traffic generation -> probabilistic)
% - Performant (verification time that makes sense, proper-level of abstraction)
% - Practical (reasonably accurate while requiring minimal amount of information and assumption)

### 2023-02-22 paper/3_functional.tex
[Paper] Updated figures and polish

\begin{figure}[h]
    \centering
    % \includegraphics[width=0.4\textwidth, trim=0cm 11cm 0cm 5cm]{ex.pdf}
    \includegraphics{../tikz/ex}
    % \includegraphics{ex.eps}
    \caption{Example topology, we want to check the reachability of A to Z}
    \label{fig:ex}
\end{figure}

@@

To represent random failure, NetDice defined a universal link failure rate (i.e. 
the probability of \textit{any} link in the network randomly failing).
We refine NetDice's model slightly by allowing each links to have different failure rate.
We label each edge in $E_t$ with a function $r: E_t \rightarrow \{x \in \mathbb{R} \mid 0 \le x 
\le 1\}$ that represents the failure rate of a given physical link.
As a consequence, two symmetrical edges (i.e. two edges that shares the same node pairs but with 
opposite direction) will also share the same failure rate, and will be disabled in a coupled 
fashion.

@@

In our running example, we could see the subset of the exploration tree in Fig. 
\ref{fig:tree}. 
We see that in a perfect network ($\mathcal{E}_1$), OSPF will produce $ACZ$ as the 
shortest path between A to Z.  
The probability of this Equivalence Class actually materializing ($P_f$) is $0.81$, 
which is the probability of $AC$ and $CZ$ being up at the same time ($0.9^2$).

In the Equivalence Class where the link $AC$ is failing however ($\mathcal{E}_2$), 
OSPF and ECMP will produce two equaly-weighted paths $ABEZ$ and $ADFZ$.
The probability of this Equivalence Class actually materializing ($P_f$) is $0.053$
($0.9^6 \cdot 0.1$).

>>>>

\begin{figure}[h]
    \centering
    % \includegraphics[width=0.4\textwidth, trim=0cm 11cm 0cm 5cm]{ex.pdf}
    \includegraphics[width=\columnwidth]{ex_topology}
    % \includegraphics{ex.eps}
    \caption{Example topology, we want to check the reachability of S to T}
    \label{fig:ex}
\end{figure}

@@

To represent random failure, NetDice defined a universal link failure rate (i.e. 
the probability of \textit{any} link in the network randomly failing).
We refine NetDice's model slightly by allowing each links to have different failure rate.
We label each edge in $E_t$ with a function 
$$r: E_t \rightarrow \{x \in \mathbb{R} \mid 0 \le x \le 1\}$$ 
that represents the failure rate of a given physical link.
As a consequence, two symmetrical edges (i.e. two edges that shares the same node pairs but with 
opposite direction) will also share the same failure rate, and will be disabled in a coupled 
fashion.

@@

In our running example, we could see the subset of the exploration tree in Fig. 
\ref{fig:tree}. 
We see that in a perfect network ($\mathcal{E}_1$), OSPF will produce $SAT$ as the 
shortest path between $S$ to $T$.  
The probability of this Equivalence Class actually materializing ($P_f$) is $0.81$, 
which is the probability of $SA$ and $AT$ being up at the same time ($0.9^2$).

In the Equivalence Class where the link $SA$ is failing however ($\mathcal{E}_2$), 
OSPF and ECMP will produce two equaly-weighted paths $SBCT$ and $SDET$.
The probability of this Equivalence Class actually materializing ($P_f$) is $0.053$
($0.9^6 \cdot 0.1$).

### 2023-02-22 paper/4_temporal.tex
[Paper] Updated figures and polish

In our Fig. \ref{fig:tree} Exploration Tree for example, $\mathcal{E}_2$ has two 
equally probable path, $ABEZ$ and $ADFZ$.
We will first do a chain convolution of all the distributions in $ABEZ$ ($l_p(AB)$, 
$l_q(AB)$, ..., $l_q(EZ)$) and do the same thing for $ADFZ$.
We will then take the $CDF$ of $\mathcal{L}_{ABEZ}$ and $\mathcal{L}_{ADFZ}$ with our 
desired $t$ and average them out to get $P_t$ of $\mathcal{E}_2$.

>>>>

In our Fig. \ref{fig:tree} Exploration Tree for example, $\mathcal{E}_2$ has two 
equally probable path, $SBCT$ and $SDET$.
We will first do a chain convolution of all the distributions in $SBCT$ ($l_p(SB)$, 
$l_q(SB)$, ..., $l_q(CT)$) and do the same thing for $SDET$.
We will then take the $CDF$ of $\mathcal{L}_{SBCT}$ and $\mathcal{L}_{SDET}$ with our 
desired $t$ and average them out to get $P_t$ of $\mathcal{E}_2$.

### 2023-02-22 paper/5_opt.tex
[Paper] updated various things
[Paper] Updated figures and polish

In our running example, we could see that in Fig. \ref{fig:tree}, $\mathcal{E}_2$ and $\mathcal{E}_4$ share 
an identical path, $ADFZ$.
We could therefore memoize $cdf(\mathcal{L}_{ADFZ}, t)$ when we explore $\mathcal{E}_2$ and query the 
cached result when we explore $\mathcal{E}_4$.

With this optimization, we will do a chain convolution by the amount of path available in the network between 
the source and destination (which is 3 in the case of our running example in Fig. \ref{fig:ex}: $ACZ$, $ABEF$, 
and $ADFZ$).

@@

% \textbf{Second}, we compute the total path latency distribution from the convergent path(s) in each 
% state with $G_l$.
% This is done by convolving over the latency distribution of each component in the path. 
% Since not all convolutions can be calculated analytically, we implemented a numerical convolution 
% method via mixture distribution, DIRECT \cite{rover2017discrete}, which is able to convolve two 
% distribution with a KL divergence error bound.
% Multiple states can share the same convergent path(s), so \textit{only a fraction of the states need to be 
% explored}.

>>>>

In our running example, we could see that in Fig. \ref{fig:tree}, $\mathcal{E}_2$ and $\mathcal{E}_4$ share 
an identical path, $SDET$.
We could therefore memoize $cdf(\mathcal{L}_{SDET}, t)$ when we explore $\mathcal{E}_2$ and query the 
cached result when we explore $\mathcal{E}_4$.

With this optimization, we will do a chain convolution by the amount of path available in the network between 
the source and destination (which is 3 in the case of our running example in Fig. \ref{fig:ex}: $SAT$, $SBCT$, 
and $SDET$).

@@

% \textbf{Second}, we compute the total path latency distribution from the convergent path(s) in each 
% state with $G_l$.
% This is done by convolving over the latency distribution of each component in the path. 
% Since not all convolutions can be calculated analytically, we implemented a numerical convolution 
% method via mixture distribution, DIRECT \cite{direct}, which is able to convolve two 
% distribution with a KL divergence error bound.
% Multiple states can share the same convergent path(s), so \textit{only a fraction of the states need to be 
% explored}.

### 2023-02-22 paper/6_eval.tex
[Paper] updated graph for fattree
[Paper] Updated figures and polish

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.5]{optimization}
    \caption{Amount of convolution (in log scale) depending on what optimization strategy is applied}
    \label{fig:opt}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[scale=0.5]{optimization_fattree}
    \caption{Amount of convolution (in log scale) depending on what optimization strategy is applied}
    \label{fig:opt_fat}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.5]{ec}
    \caption{Amount of temporal EC compared to functional EC, label represents ratio}
    \label{fig:ec}
\end{figure}

@@

The first one is the \textbf{amount of possible paths}.
If the amount of possible paths between a src-dst pair in a given topology is small, then the functional 
verification step would produce fewer equivalence classes and end early.
While most of our evaluated topology had a lot of possible paths, one exception to this is LatNet, which 
only have 3 possible paths.

The second is whether the src-dst pair in a given topology would \textbf{produce paths that are "symmetric"}.
By symmetric, we mean that the routing and load balancing method produce a lot of identical equivalence class 
and / or equivalence class that share the same path.
As our explanation about Fat Tree 10 suggests, larger Fat Tree has a lot of path with the same length.
This would result in a lot of symmetrical equivalence classes under ECMP.
Our optimization techniques could identify this symmetry and reduce the amount of convolution in the temporal 
verification step.

>>>>

\begin{figure}[h]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=\textwidth]{optimization}
        \caption{Amount of convolution (in log scale) depending on what optimization strategy is applied}
        \label{fig:opt}
    \end{subfigure}
    \vfill
    \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=\textwidth]{optimization_fattree}
        \caption{Amount of convolution (in log scale) depending on what optimization strategy is applied. 
            Fattree 12 upwards is extrapolatied from the number of convolution}
        \label{fig:opt_fat}
    \end{subfigure}
    \caption{test}
\end{figure}

@@

% \begin{figure}[h]
%     \centering
%     \includegraphics[scale=0.5]{ec}
%     \caption{Amount of temporal EC compared to functional EC, label represents ratio}
%     \label{fig:ec}
% \end{figure}

\begin{figure}[h]
    \centering
    \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=\textwidth]{ec}
        \caption{Functional EC and Temporal EC for WAN topologies}
        \label{fig:y equals x}
    \end{subfigure}
    \vfill
    \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=\textwidth]{ec_fattree}
        \caption{Functional EC and Temporal EC for DC topologies}
        \label{fig:three sin x}
    \end{subfigure}
       \caption{Amount of temporal EC compared to functional EC, label represents ratio}
       \label{fig:ec}
\end{figure}

@@

The first one is the \textbf{amount of possible paths} (Fig. \ref{fig:opt_analysis_amt}).
If the amount of possible paths between a src-dst pair in a given topology is small, then the functional 
verification step would produce fewer equivalence classes and end early.
In our simplified example in Fig. \label{fig:opt_analysis_amt}, although the network has multiple interconnected
nodes, S and T has only one possible path.
While most of our evaluated topology had a lot of possible paths, one exception to this is LatNet, which 
only have 3 possible paths in our evaluation.

The second is whether the src-dst pair in a given topology would \textbf{produce paths that are "symmetric"}
(Fig \ref{fig:opt_analysis_symmetry}).
By symmetric, we mean that the routing and load balancing method produce a lot of identical equivalence class 
and / or equivalence class that share the same path.
In our simplified example in Fig. \ref{fig:opt_analysis_symmetry}, the network has five 2-hop paths, which will give a 
lot of equal-cost path combination under ECMP.
As our explanation about Fat Tree 10 suggests, larger Fat Tree has a lot of path with the same length.
This would result in a lot of symmetrical equivalence classes under ECMP.
Our optimization techniques could identify this symmetry and reduce the amount of convolution in the temporal 
verification step.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{eval_ex_amt}
        \caption{S-T only has one path}
        \label{fig:opt_analysis_amt}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{eval_ex_symmetry}
        \caption{S-T has five 2-hop paths}
        \label{fig:opt_analysis_symmetry}
    \end{subfigure}
       \caption{Illustration of topological property}
       \label{fig:opt_analysis}
\end{figure}
